---
title: "Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Ryan McGuinness


#### Importing Weather Data

Weather data for Cleveland on the day of Sept. 10, 2021 was collected from [Weather Underground](https://www.wunderground.com/history/daily/us/oh/cleveland/KCLE/date/2021-9-10). Each record includes time recorded, temperature, wind characteristics, and more. 

```{r}
weather_data <- read.csv("weather.csv")
weather_data
```



#### Adding Collumn to Indicate Time Elapsed Since Midnight

In order to plot Temperature, we need a quantitative time variable. Let's use minutes passed since midnight as that variable.
```{r}
weather_data$hour <- strtoi(c(data.frame(strsplit(weather_data$Time, ":"))[1,]))
weather_data$isPM <- c(data.frame(strsplit(weather_data$Time, " "))[2,]) == "PM"

weather_data$Time_Since_Midnight <- c(60*(weather_data$hour + 12*(weather_data$isPM - (weather_data$hour == 12))) + 51) 
# Each record is taken 51 minutes past the hour
```

Now, let's check that time variable.
```{r}
weather_data[c("Time", "Time_Since_Midnight", "Temperature")]
```


#### Including Plot for Temperature vs Time

```{r}
plot(weather_data$Time_Since_Midnight, weather_data$Temperature,
     xlab="Time Since Midnight (min)", ylab="Temperature (F)")
```


