---
title: "Assignment 2"
output: html_document
author: "Ryan McGuinness"
---

```{r setup, include=FALSE, echo=FLASE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Required packages
library(class)
library(caret)
```


## Reading data
Data for training, validation, and testing is included in `UniversalBank.csv`. The first step of the project is to read that data. Education is a dummy variable with three possible values, so that will need to be broken into three variables.


```{r}
# Reading bank data
Bank.Data <- read.csv("UniversalBank.csv")

# Splitting Education into three variables
Bank.Data$Education_1 <- Bank.Data$Education == 1
Bank.Data$Education_2 <- Bank.Data$Education == 2
Bank.Data$Education_3 <- Bank.Data$Education == 3

head(Bank.Data)
```

ID and ZIP Code need not be used by the model, so we will remove them. Also, we can remove the old Education variable since we broke it into three components.

```{r}
# Removing ID (1), ZIP (5), and Education (8)
Bank.Selected <- Bank.Data[c(-1, -5, -8)]

head(Bank.Selected)
summary(Bank.Selected)
```


## Data splitting
Next, we divide the data into training and validation. For questions 1-4, we need 60% of our data to be training, and 40% validation.

```{r}
# Setting random seed to get reproducible results
set.seed(2468)

# Splitting data into training and validation
Train.Index <- createDataPartition(Bank.Selected$Personal.Loan, p=0.6, list=FALSE)
Train.Data <- Bank.Selected[Train.Index,]
Valid.Data <- Bank.Selected[-Train.Index,]
```

## Normalizing
We normalize the numeric predictors for training and validation data.

```{r}
# Copying original data
Train.Normalized <- Train.Data
Valid.Normalized <- Valid.Data

# Creating normalization model for first 6 variables in training data
norm.model <- preProcess(Train.Data[,1:6], method=c("center", "scale"))
Train.Normalized[,1:6] <- predict(norm.model, Train.Data[,1:6])
Valid.Normalized[,1:6] <- predict(norm.model, Valid.Data[,1:6])

summary(Train.Normalized)
summary(Valid.Normalized)
```

## Modelling k-NN
For questions 1 and 4, we can given a particular customer.

```{r}
# Creating data for a customer as described by question 1
Customer <- data.frame(Age=40, Experience=10, Income=84, Family=2, CCAvg=2,
                       Mortgage=0, Securities.Account=0, CD.Account=0, Online=1,
                       CreditCard=1, Education_1=FALSE, Education_2=TRUE, 
                       Education_3=FALSE)

# Normalizing the customer
Customer.Normalized <- Customer
Customer.Normalized[,1:6] <- predict(norm.model, Customer[,1:6])
```

We use the `knn()` function to find the single nearest neighbor to the customer to predict whether the customer will accept the personal loan.
```{r}
nn <- knn(Train.Normalized[,-7], Customer.Normalized, Train.Normalized[,7],
          k=1, prob=TRUE)

nn[1]
```
The k-NN model with k=1 predicts this customer will not accept a personal loan.

## Finding the best k
We will let k vary from 1 to 10 to see which choice of k is the best fit for this model. 
```{r}
max.k <- 10
knn.accuracy <- data.frame(k=(1:max.k), accuracy=rep(0, times=max.k))

actual <- factor(Valid.Normalized[, 7])

for (i in 1:max.k){
  predicted <- knn(Train.Normalized[,-7], Valid.Normalized[,-7],
                   Train.Normalized[,7], k=i)
  knn.accuracy[i, 2] <- confusionMatrix(predicted, actual)$overall[1]
}

knn.accuracy
```

It appears our best choice of k is k=3.
