---
title: "Assignment 3"
author: "Ryan McGuinness"
date: "10/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Description
We use the Naive Bayes classification model to ultimately compute the proportion of loan accepters among online banking users with credit cards issued by the bank in the `UniversalBank.csv` dataset. 

## Data Preparation

First, we load required libraries.
```{r}
library(e1071)
library(caret)
```

We read the data from `UniversalBank.csv`.
```{r}
Bank.Data <- read.csv("UniversalBank.csv")
head(Bank.Data)
```

Next we must transform the data set. We only care about Personal.Loan, Online, and CreditCard variables. These variables then need to be transformed into factors. 
```{r}
# Removing all irrelevant variables, leaving online Personal.Loan, Online, and CreditCard
Bank.Data <- Bank.Data[,c(10, 13, 14)]

# Converting categorical variables into factors
Bank.Data$Personal.Loan <- factor(Bank.Data$Personal.Loan)
Bank.Data$Online <- factor(Bank.Data$Online)
Bank.Data$CreditCard <- factor(Bank.Data$CreditCard)

summary(Bank.Data)
```

The data is partitioned into training set (60%) and validation set (40%).
```{r}
# Setting random seed
set.seed(127429)

# Dividing Bank.Data into training and validation sets 60%/40%
Train.Index <- createDataPartition(Bank.Data$Personal.Loan, p=0.6, list=FALSE)
Train.Data <- Bank.Data[Train.Index,]
Valid.Data <- Bank.Data[-Train.Index,]
```

## Part A: Visualizing Training Set
The following table conveys the counts in the training set where the CC and Online variables are considered together.
```{r}
Online.CC.Table <- table(Train.Data$Personal.Loan, Train.Data$Online, Train.Data$CreditCard, dnn=c("Personal Loan", "Online", "Credit Card"))
Online.CC.Table
```

## Part B: Computing P(Loan | Online & CC) Directly
We can look at the table from Part A to get an immediate calculation of the proportion of loan accepters among online banking users with credit cards issued by the bank. Of the 537 records where of active online banking users with credit cards, 48 had accepted a personal loan, so $\textbf{P}(\text{Loan}=1\ |\ \text{CC}=1,\text{ Online}=1) = \frac{48}{537} = 0.089$.
```{r}
# Computing P(loan | Online & CC)
Online.CC.Table[2,2,2] / (Online.CC.Table[2,2,2] + Online.CC.Table[1,2,2])
```


## Part C: Visualizing Separated Training Set 
The following tables convey the counts in the training set for the CC and Online variables separately.
```{r}
Online.Table <- table(Train.Data$Personal.Loan, Train.Data$Online, dnn=c("Personal Loan", "Online"))
Online.Table
CC.Table <- table(Train.Data$Personal.Loan, Train.Data$CreditCard, dnn=c("Personal Loan", "Credit Card"))
CC.Table
```

## Part D: Computing various probabilities for Naive Bayes method
In order to compute $\textbf{P}\text{(Loan = 1 | Online = 1 & CC = 1)}$, we need to find the following six probabilities: $\textbf{P}\text{(CC = 1 | Loan = 1)}$, $\textbf{P}\text{(Online = 1 | Loan = 1)}$, $\textbf{P}\text{(Loan = 1)}$, $\textbf{P}\text{(CC = 1 | Loan = 0)}$, $\textbf{P}\text{(Online = 1 | Loan = 0)}$, and $\textbf{P}\text{(Loan = 0)}$.
```{r}
# P(Online | Loan)
p_ol <- Online.Table[2,2] / (Online.Table[2,2] + Online.Table[2,1])
p_ol
# P(CC | Loan)
p_cl <- CC.Table[2,2] / (CC.Table[2,2] + CC.Table[2,1])
p_cl
# P(Loan)
p_l <- sum(Train.Data$Personal.Loan==1) / length(Train.Data$Personal.Loan)
p_l
# P(Online | !Loan)
p_onl <- Online.Table[1,2] / (Online.Table[1,2] + Online.Table[1,1])
p_onl
# P(CC | !Loan)
p_cnl <- CC.Table[1,2] / (CC.Table[1,2] + CC.Table[1,1])
p_cnl
# P(!Loan)
p_nl <- sum(Train.Data$Personal.Loan==0) / length(Train.Data$Personal.Loan)
p_nl
```

## Part E: Computing P(Loan | Online & CC) Via Naive Bayes
Using the Naive Bayes method and the probabilities computed in Part D, we find $\textbf{P}(\text{Loan}=1\ |\ \text{CC}=1,\text{ Online}=1) = 0.096$.
```{r}
(p_ol * p_cl * p_l) / (p_ol * p_cl * p_l + p_onl * p_cnl * p_nl)
```

## Part F: Comparing Naive Bayes to Direct Calculaton
Using the Naive Bayes classifier, we get a higher value for $\textbf{P}(\text{Loan}=1\ |\ \text{CC}=1,\text{ Online}=1)$ than with the direct computation. Interestingly, the Naive Bayes approach gives $\textbf{P}(\text{Loan}=1\ |\ \text{CC}=1,\text{ Online}=1) = 0.096$, but in Part D we found that $\textbf{P}\text{(Loan = 1)} = 0.096$ to begin with. So the Naive Bayes approach suggests that the probability a person will accept the loan is independent of whether that person is an online user with a bank-issued credit card.

